---
title: "Scoring fly enhancers using NRLB models"
author: "Chaitanya Rastogi and Harmen J. Bussemaker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring fly enhancers using NRLB models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette defines a set of fly enhancers, loads DNA binding specificity models for various Hox complexes, and then visualizes Hox binding affinity profiles along the enhancers. It is related to the paper on the No Read Left Behind (NRLB) algorithm by Rastogi et al., which is currently in revision.

## Defining fly enhancers

We start by defining a number of fly enhancers that were used in the NRLB paper. We will only be using the E3N in this vignette, but the others are included for reference. Note that the version of the fly genome used here is the somewhat outdated `dm3`. The coordinates are only 'guaranteed' to be correct for `dm3`.
```{r}
enh = NRLBtools::fly.enhancers()
names(enh)
enh$E3N
```
## DNA binding specificity models
Next, we define and load a number of relevant DNA binding models that were built from SELEX-seq in vitro binding data using the NRLB algorithm. These binding models are included with the current package. They are identical to what was used in the NRLB paper.
```{r}
models = NRLBtools::hox.models()
names(models)
```
## Energy Logo
Let us select a model for a specific heterodimer from our model collection:
```{r}
m = models$ExdScr
```
It is always a good idea to visualize the model as an energy logo before using it:
```{r}
NRLBtools::logo(m$fits, m$index, m$mode)
```

## Affinity profiles
Finally, we can created an affinity profile along the E3N enhancer as follows:
```{r, fig.show='asis'}
model = models$ExdScr
NRLBtools::nrlb.plot.score.genome(enh$E3N$seq,
                                  m$fits, m$index, m$mode,
                                  nPeaks = 10, annotate = TRUE)
```